A missing guardrail: The failure of protest risk to deter election manipulation in non-democracies

### 2023-06-05 Output/Submission files/manuscript_anon_v2.Rmd
Updates to text + section on Cox models

As a final test of the mechanism outlined in the theory presented here\textemdash that election manipulation offers few clues to regime strength, and thus the course of electoral protests are determined by more structural factors\textemdash the same models will also be applied to the duration of post-election protest waves. Longer protest waves indicate a more even match between the resources and resiliency of the protesters and the capacity and cohesiveness of the regime. It is expected that more severe election fraud will *not* be associated with longer protest waves, as would be predicted if fraud were an indicator of state weakness. Instead, structural factors should again be more influential;   harsh repression should be associated with shorter protest waves, while indicators of opposition and civil-society strength should help keep protest waves alive.^[Note here about excluding successful movements.]

@@

nelda2.sub <- nelda2.sub %>% mutate(presidential = factor(presidential))
nelda2.sub <- nelda2.sub %>% mutate(inc.termlimit.num = factor(inc.termlimit.num, exclude = NULL))
nelda2.sub <- nelda2.sub %>% mutate(rentierstate = factor(rentierstate))
nelda2.sub <- nelda2.sub %>% mutate(taxstate = factor(taxstate))
nelda2.sub <- nelda2.sub %>% mutate(unscheduled_election = factor(unscheduled_election))
nelda2.sub <- nelda2.sub %>% mutate(inc.lose = factor(inc.lose))
nelda2.sub <- nelda2.sub %>% mutate(econ.crisis.nelda.1lag = factor(econ.crisis.nelda.1lag))

@@

These hypotheses are tested using data from three datasets: NELDA [@hyde2012], V-Dem version 11.1 [@vdem11.1], and the Electoral Contention and Violence dataset [@Daxecker2019]. The unit of observation for this study is an election period as recorded in NELDA. This framework prevents selecting on the dependent variable, by capturing both protested and non-protested elections.  I limit the sample by excluding closed autocracies (where no national elections are held). Since the theory of backlash to electoral manipulation could in theory apply to liberal democracies, I include them in the main analysis (as coded by V-Dem).^[The specific variable used is 'v2x_regime'.] As a robustness check for this sampling decision, I also include the main models without liberal democracies in the appendix, with no substantial changes in the outcomes. Finally, the beginning of the date range is limited by the availability of data on protest from ECAV, as discussed below, which ranges from 1990 to 2012. The resulting dataset includes 647 total election-period observations across 109 countries.

## Dependent variables

To capture the *severity of election fraud*, I use the V-Dem measure of intentional election irregularities, a category which includes "double IDs, intentional lack of voting materials, ballot-stuffing, misreporting of votes, and false collation of votes." I operationalize electoral manipulation using this variable, rather than aggregate measures of election integrity, because fraud of this kind is thought to be most likely to trigger protest [@Linebarger2020]---it is often incumbents' tool of last resort [@Sjoberg2016], it can fail to signal strength [@Harvey2020], and it can be more damaging to legitimacy than pre-election manipulation [@birch2011;@Szakonyi2021]. This measure of election integrity should thus serve as a strong test of the protest-oriented model. The variable is used as a dependent variable in the first set of models, and as a potential explanatory variable in the subsequent models of protest number, size, and duration.

ECAV records protests and other contentious actions related to an election, ranging from six months before to three months after the vote (or three months after the final round of a multi-round election). To focus the study on protest, I only include events labeled as protests, occupations, or blockades, excluding acts like shootings, bombings, and coups. ECAV data is the event-day format; to capture the *number of protests* before and after the election, I take the sum number of events recorded in ECAV before and after election day, with election-day protests being counted as post-election. To evaluate hypotheses regarding protest size, I convert ECAV's categorical data on number of participants per event-day to estimate numbers, and take the sum where multiple protests were held on a given day. The ECAV dataset codes protest size as a categorical variable with five levels, ranging from fewer than ten participants to more than 10,000. To deal with this data structure, where the top category has no upper limit, I take the median value of each category, with the exception of the top category. For this category, which has no upper bound, I take the minimum value of 10,000. The variable *maximum protest size* is found by taking the number of estimated protest participants who turned out on a single day.

The large majority of elections experience no post-election protest: in 1168 of the 1477 elections, no protests are recorded. Figure \@ref(fig:elex-hist) reports the distribution of election fraud across the dataset, as well as protest number and maximum single-day protest size.

@@

```{r, include=F}
incl.vars <- nelda2.sub %>% dplyr::select('Judicial independence' = jucon.1lag, 
                                            'Legislative constraints' = legcon.1lag, 
                                            'Alternative information' = altinf.1lag, 
                                            'GDP growth rate' = gdpgro.1lag, 
                                            'Civil society openness' = osorg.1lag, 
                                            'Public sector corruption' = pubseccorrup.lag,
                                            'Economic crisis' = econ.crisis.nelda.1lag,
                                            'National party organization' = natlpartyorg.lag, 
                                             'Election fraud' = v2elirreg.inv, 
                                            'Presidential election' = presidential,
                                            'Incumbent term-limited' = inc.termlimit.num,
                                            'Physical integrity index' = physinteg.1lag,
                                            'Urbanization' = urban.pct, 
                                            'Rentier state' = rentierstate,
                                            'Tax state' =  taxstate,
                                            'Incumbent vote-share' = winner.share.upd, 
                                            'Median pre-election protest size' = median.participants.pre,
                                            'Largest protest size' = max_size,
                                            'Off-schedule election' = unscheduled_election)
#incl.vars <- incl.vars %>% mutate(v2elintmon = as.numeric(v2elintmon))

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(kableExtra)
datasummary_skim(incl.vars, output = "kableExtra")  %>%
    kable_styling(font_size = 9, full_width = FALSE)
```
\doublespace

This distribution\textemdash with large numbers of zeroes\textemdash is in line with the theoretical proposition that two data-generating processes are at work: the factors that lead to the occurrence of protest may not be the same as those that sustain or intensify it. Consequently, I model the results using negative binomial hurdle regressions. Hurdle models are two-part regressions in which the variation in the dependent variable is assumed to follow two distributions: the probability of obtaining a zero or non-zero outcome is modeled as binomial logic function, while the distribution of the non-zero count variable is modeled (in this case) using the negative binomial function. The logistic portion of the model identifies the factors that make *any protest* more likely to occur; the negative binomial regressions to model the count of protests given that protest has begun. Since protest size cannot take on a negative value, it is also modeled as a count variable with a negative binomial distribution in the hurdle models.

@@

The first set of models test Hypothesis 1, on the relationship between protest risk and fraud. Protest risk is operationalized in three ways. First, the *number of pre-election protests* in ECAV is taken as *prima facie* evidence of a more turbulent and risky election environment\textemdash it indicates greater mobilization capacity in civil society and/or the opposition, and may also indicate an elevated sense of grievance. Second, the repressive capacity of the regime is modeled using the lagged *physical integrity index* for the country-year. In one approach, higher values represent more permissive environments, and should be associated with greater protest risk (and thus reduced fraud). Alternatively, in keeping with the 'murder in the middle' framework, the highest protest risks (and most reduced fraud) might be found at middle values of repression. Finally, the literature on the diffusion of electoral protest holds that successful electoral protest in one country increases the likelihood of electoral protest in similar countries nearby. This diffusion mechanism is modeled using each country's geopolitical region as coded in Teorell et al [-@qog_teorell] in conjunction with binary variables indicating a successful protest movement from NELDA. A 'successful protest' is recorded if a protest movement resulted in either a new election being run or the incumbent stepping down. The *regional diffusion* dummy variable takes on a value of one for all countries in the same region as a successful protest, for elections within one calendar of the successfully overturned election.^[For example, the variable takes on a positive value for elections held in the post-communist region, within one year of the September 24, 2000 election in Serbia that sparked the 'Bulldozer Revolution'. All told there are xx successful electoral protests in the dataset: India 1989, Bulgaria 1990, Bangladesh 1996, Mali 1997, Lesotho 1998, Cote d'Ivoire 2000 and 2010, Peru 2000, Serbia 2000, Georgia 2003, Ukraine 2003-2004, Azerbaijan 2005, Kyrgyzstan 2005, Haiti 2006, and Cameroon 2007]

@@

I measure repressive capacity using the *physical integrity* index from V-Dem, lagged one year. Higher values for this variable indicate greater freedom from torture and political killing. I also include the variable's square term to account for the 'murder in the middle' framework. The physical integrity variables are expected to be explanatory in models of protest, but are important control variables in the models of election fraud. Since repressive capacity helps influence protest in expectation [@Ritter2016], incumbents that are more confident of deterring protest in advance should be more willing to engage in fraud under the protest-deterrent model of manipulation.

Here I expect to see support for the so-called 'more murder in the middle' framework [@feinLifeIntegrityViolationsDemocracy1995], in which the regimes with the most demonstrated repressive capacity deter potential challengers from taking to the streets in the first place, and the least repressive states defuse protest through accommodation [@pierskallaProtestDeterrenceEscalation2010;@mooreRepressionDissentSubstitution2000].

@@

Several control variables are included, which could be common causes for both the severity of election manipulation and protest. To help control for the underlying mobilizational capacity, which likely affects incumbents' decision to manipulate and the latent risk of protest, I include lagged indicators of government *control over civil society* (the 'v2cseeorgs' variable from V-Dem), the proportion of national political parties with *permanent party organizations* (V-Dem's 'v2psorgs'), and the available of *alternative sources of information*. Since urban areas can make electoral manipulation more difficult [@ziblatt2009;@larreguy2016parties] while also making protest more likely [@Wallace2013a], I also include data on *urbanization* taken from the United Nations [@DepartmentofEconomicandSocialAffairsPopulationDivision2019]. Finally, all models include measures of *judicial constraints* and  *legislative constraints* on the executive, V-Dem variables which captures the extent to which the judicial and legislative bodies in the country act independently in practice to limit executive power.

Finally, several binary control variables are included. First, a variable from NELDA is used to indicate *presidential* elections, since these may include both higher levels of manipulation [@Simpser2011] and a greater risk of protest. Next, a variable from NELDA indicates if the chief executive is *term limited* during the election at hand. This variable is included since incumbents running up against constitutional term limits can pose major challenges for the unity of the regime, possibly leading to fractured manipulation efforts and post-election protest [@hale_patronal_2014]. Finally, I include dummy variables that provide a categorical evaluation of state capacity. These are taken from the ordinal version of the V-Dem variable "state source of fiscal revenue" (v2stfisccap). If the V-Dem data codes a state as not capable of raising revenue or primarily relying on loans and foreign aid, I code it as a *weak state*. If the state is described as primarily relying on resource rents and expropriation, I code it is a *rentier state*. States that rely primarily on taxes on property or economic transactions are coded as *tax states*. This distinction is relevant since access to natural resource rents may both make protest less likely and repression more effective [@Girod2018]. Lastly, since strong non-democratic governments may be able to adjust the electoral calendar strategically (to affecting both election manipulation and protest risk), I include a variable from NELDA that indicates *off-schedule elections*. Figure 2 shows summary statistics for all of these variables. Lastly, a binary variable from NELDA indicates elections in which *incumbents lost*; I expect that protest will be less likely when this variable takes a value of one.  To address concerns about multicollinearity, I present models in the appendix (page 25) that exclude control variables associated with pre-election protest.

@@

The results of these models are not promising for argument that protest risk deters election manipulation. First, Hypothesis 1 is not supported in Models 1 and 2 in Table  \@ref(tab:manipdv) \textemdash the coefficient on *pre-election protest* is positive and significant. The model predicts that each additional pre-election protest event is associated with an **increase** in the fraud measure of 0.008. This is a very small effect, given that the dependent variable ranges from `r min(nelda2.sub$v2elirreg.inv, na.rm = T)` to `r max(nelda2.sub$v2elirreg.inv, na.rm = T)`. An increase from zero to five protest events results in a predicted increase in fraud of roughly three percent of a standard deviation. While this is a small effect, it runs counter to the expectations of the deterrence logic. And the effect may be sizable in extreme circumstances. For example, pre-election protest in the 99th percentile is associated with an increase of `r (as.numeric(quantile(nelda2.sub$n.events.pre, probs = .99)))*.008`, about one-fifth of a standard deviation. This suggests that the level of election manipulation, rather than being deterred by a highly contentious environment, may be deployed to make up for anticipated poor election results.

@@

```{r manipdv, echo=FALSE}
m1 <- lmer(v2elirreg.inv ~ jucon.1lag + legcon.1lag + gdpgro.1lag +  osorg.1lag  + natlpartyorg.lag   +  presidential + libdem.1lag + urban.pct + rentierstate + taxstate + n.events.pre  +  unscheduled_election  + (1 | stateid), data = nelda2.sub)

@@

m1a <- lmer(v2elirreg.inv ~ jucon.1lag + legcon.1lag + gdpgro.1lag +  osorg.1lag  + natlpartyorg.lag   +  presidential + libdem.1lag + urban.pct + rentierstate + taxstate +  unscheduled_election + regional_success_lag + (1 | stateid), data = nelda2.sub) 
#summary(m1a)

m1b <- lmer(v2elirreg.inv ~ jucon.1lag + legcon.1lag  + gdpgro.1lag +  osorg.1lag  + natlpartyorg.lag   + presidential   +  libdem.1lag + urban.pct + rentierstate + taxstate + physinteg.1lag +  unscheduled_election + (1 | stateid), data = nelda2.sub)

@@

m1c <- lmer(v2elirreg.inv ~ jucon.1lag + legcon.1lag + gdpgro.1lag +  osorg.1lag  + natlpartyorg.lag   +  presidential + libdem.1lag + urban.pct + rentierstate + taxstate  + physinteg.1lag + physintlag.sq + unscheduled_election  + (1 | stateid), data = nelda2.sub) 
#summary(m1c)

modelsummary(c(m1, m1b, m1c, m1a), output = "kableExtra",  stars = T, gof_omit = c("R2|RMSE"), title = 'Random-intercept multilevel model of election-day fraud', coef_rename = 
               c("jucon.1lag" = "Judical independence (lag)", "legcon.1lag" = "Leg. constraints (lag)", 
                 "gdpgro.1lag" = "GDP growth rate (lag)", "altinf.1lag" = "Alternative info. (lag)", 
                 "osorg.1lag" = "Civil soc. openness (lag)", "econ.crisis.nelda.1lag1" = "Economic crisis", 
                 "natlpartyorg.lag" = "National party organization (lag)", "presidential1" = "Presidential election",
                 "inc.termlimit.num" = "Incumbent term-limited", "libdem.1lag" = "Liberal democracy index (lag)", 
                 "urban.pct" = "Urbanization", "rentierstate" = "Rentier state", "taxstate" = "Tax state",
                 "n.events.massmob.pre" = "Number of pre-election protests", "median.participants.mm.pre" = "Median size of pre-election protests", "unscheduled_election" = "Off-schedule election")) %>% kable_styling(latex_options = "HOLD_position")

@@

```{r models-hurdle-mm, warning = FALSE, message = FALSE, include = F}
model.n.novoteshare.hurd.mm <- glmmTMB::glmmTMB(n.events.post ~  jucon.1lag + legcon.1lag + altinf.1lag + gdpgro.1lag + pubseccorrup.lag + osorg.1lag  + natlpartyorg.lag + v2elirreg.inv + pubseccorrup.lag +    + v2elirreg.inv  + presidential + inc.termlimit.num  +  physinteg.1lag + physintlag.sq + urban.pct + rentierstate + taxstate + n.events.pre + unscheduled_election  + regional_success_lag,
                                ziformula=~.,
                                data = nelda2.sub %>% filter(inc.lose == "0"),
                                family=truncated_nbinom2)
summary(model.n.novoteshare.hurd.mm)

model.n.base.hurd.mm <- glmmTMB(n.events.post ~ jucon.1lag + legcon.1lag + altinf.1lag + gdpgro.1lag + pubseccorrup.lag + osorg.1lag  + natlpartyorg.lag + v2elirreg.inv + pubseccorrup.lag +    winner.share.upd + v2elirreg.inv  + presidential + inc.termlimit.num  +  physinteg.1lag + physintlag.sq + urban.pct + rentierstate + taxstate + n.events.pre + unscheduled_election + regional_success_lag,
                               ziformula=~.,
                              data = nelda2.sub %>% filter(inc.lose == "0"),
                                family=truncated_nbinom2)
summary(model.n.base.hurd.mm)

model.n.base2.hurd.mm <- glmmTMB::glmmTMB(n.events.post ~  jucon.1lag + legcon.1lag + gdpgro.1lag + altinf.1lag + pubseccorrup.lag + osorg.1lag  + natlpartyorg.lag + v2elirreg.inv + pubseccorrup.lag +    winner.share.upd*v2elirreg.inv  + presidential + inc.termlimit.num  +  physinteg.1lag + physintlag.sq + urban.pct + rentierstate + taxstate + n.events.pre + unscheduled_election + regional_success_lag,
                                          zi = ~.,
                                          family= nbinom2,
                                          data = nelda2.sub %>% filter(inc.lose == "0"))
summary(model.n.base2.hurd.mm)

>>>>

To further test the mechanism outlined in the theory presented here\textemdash that election manipulation offers few clues to regime strength, and thus the course of electoral protests are determined by more structural factors\textemdash I examine the relationship between fraud and the survival of protest movements in cases where protests began but the incumbent eventually survived. Restricting the sample to these cases allows for a consistent interpretation of results: longer protest waves indicate a more even match between the resources and resiliency of the protesters and the capacity and cohesiveness of the regime, while shorter-lived protest movements are those which were more easily defeated.  In this sub-sample, I would first expect the relationship between election fraud and protest survival to be non-linear with respect to time\textemdash the longer the protest movement survives, the less relevant the fraud signal (such as it is) becomes). Second, it is expected that more severe election fraud will *not* be associated with longer protest waves, as would be predicted if fraud were an indicator of state weakness. Instead, structural factors should again be more influential; indicators of opposition and civil-society strength should help keep protest waves alive.

@@

nelda2.sub <- nelda2.sub %>% mutate(presidential = factor(presidential))
#nelda2.sub <- nelda2.sub %>% mutate(inc.termlimit.num = factor(inc.termlimit.num))
nelda2.sub <- nelda2.sub %>% mutate(rentierstate = factor(rentierstate))
nelda2.sub <- nelda2.sub %>% mutate(taxstate = factor(taxstate))
nelda2.sub <- nelda2.sub %>% mutate(unscheduled_election = factor(unscheduled_election))
nelda2.sub <- nelda2.sub %>% mutate(inc.lose = factor(inc.lose))
nelda2.sub <- nelda2.sub %>% mutate(econ.crisis.nelda.1lag = factor(econ.crisis.nelda.1lag))

@@

These hypotheses are tested using data from three datasets: NELDA [@hyde2012], V-Dem version 11.1 [@vdem11.1], and the Electoral Contention and Violence dataset [@Daxecker2019]. The unit of observation for this study is an election period as recorded in NELDA. This framework prevents selecting on the dependent variable, by capturing both protested and non-protested elections.  I limit the sample by excluding closed autocracies (where no national elections are held). Since the theory of popular backlash to manipulation has largely been articulated in an electoral-democratic context, I exclude liberal democracies from the main analysis.^[The specific variable used is 'v2x_regime'; I limit the sample to those regime-years coded 1 or 2.] As a robustness check for this sampling decision, I also run the main models with data including liberal democracies in the appendix, with no substantial changes in the outcomes. The first set of models, focusing on Hypothesis 1, include all the remaining cases in the dataset. For the subsequent models, focusing on post-election protest, I exclude cases where the incumbent party lost, based on coding in NELDA\textemdash this limits the observations to those that fit the theoretical expectations of protest risk following an election in which the incumbent won, possibly through reliance on election fraud. Finally, the beginning of the date range is limited by the availability of data on protest from ECAV, as discussed below, which ranges from 1989 to 2012. The resulting dataset includes 1110 total election-period observations across 118 countries.

## Dependent variables and modeling strategies

To capture the severity of *election fraud*, I use the V-Dem measure of intentional election irregularities, a category which includes "double IDs, intentional lack of voting materials, ballot-stuffing, misreporting of votes, and false collation of votes." I operationalize electoral manipulation using this variable, rather than aggregate measures of election integrity, because fraud of this kind is thought to be most likely to trigger protest [@Linebarger2020]---it is often incumbents' tool of last resort [@Sjoberg2016], it can fail to signal strength [@Harvey2020], and it can be more damaging to legitimacy than pre-election manipulation [@birch2011;@Szakonyi2021]. This measure of election integrity should thus serve as a strong test of the protest-oriented model. The variable is used as a dependent variable in the first set of models that test H1, and as a potential explanatory variable in the subsequent models of protest number, size, and duration.

ECAV records protests and other contentious actions related to an election, ranging from six months before to three months after the vote (or three months after the final round of a multi-round election). To focus the study on protest, I only include events labeled as protests, occupations, or blockades, excluding acts like shootings, bombings, and coups. ECAV data is the event-day format; to capture the *number of protests* before and after the election, I take the sum number of events recorded in ECAV before and after election day, with election-day protests being counted as post-election. To evaluate hypotheses regarding protest size, I convert ECAV's categorical data on number of participants per event-day to estimate numbers, and take the sum where multiple protests were held on a given day.^[This approach is necessary to avoid conflating days with, for example, five category 1 protests (totaling perhaps 50 participants) and days with a single category 5 protest (totaling more than 10,000).] The ECAV dataset codes protest size as a categorical variable with five levels, ranging from fewer than ten participants to more than 10,000. To deal with this data structure, where the top category has no upper limit, I take the median value of each category, with the exception of the top category. For this category, which has no upper bound, I take the minimum value of 10,000. The variable *maximum protest size* is found by taking the number of estimated protest participants who turned out on a single day.

The large majority of elections experience no post-election protest: in `r nelda2.sub %>% filter(n.events.post == 0) %>% nrow()` of the `r nrow(nelda2.sub)` elections, no protests are recorded. Figure \@ref(fig:elex-hist) reports the distribution of election fraud across the dataset, as well as protest number and maximum single-day protest size. As the figure shows, while election fraud is common among this set of regimes, protests are rare, and usually small in number and maximum size when they do occur. This distribution\textemdash with large numbers of zeroes\textemdash is in line with the theoretical proposition that two data-generating processes are at work: the factors that lead to the occurrence of protest may not be the same as those that sustain or intensify it. Consequently, I model the number of protests using negative binomial hurdle regressions. Hurdle models are two-part regressions in which the variation in the dependent variable is assumed to follow two distributions: the probability of obtaining a zero or non-zero outcome is modeled as binomial logit function, while the distribution of the non-zero count variable is modeled (in this case) using the negative binomial function. The logistic portion of the model identifies the factors that make *any protest* more likely to occur; the negative binomial regressions model the count of protests given that protest has begun. Since protest size cannot take on a negative value, it is also modeled as a count variable with a negative binomial distribution in the hurdle models. *Election fraud*, which follows a more normal distribution, is modeled using multilevel models with random intercepts by country.

@@

```{r, include=F}
incl.vars <- nelda2.sub %>% dplyr::select('Judicial independence' = jucon.1lag, 
                                            'Legislative constraints' = legcon.1lag, 
                                            'Alternative information' = altinf.1lag,
                                          'Liberal democracy index' = libdem.1lag,
                                            'GDP growth rate' = gdpgro.1lag, 
                                            'Civil society openness' = osorg.1lag, 
                                            'Public sector corruption' = pubseccorrup.lag,
                                            'National party organization' = natlpartyorg.lag, 
                                             'Election fraud' = v2elirreg.inv, 
                                            'Presidential election' = presidential,
                                            'Incumbent term-limited' = inc.termlimit.num,
                                            'Physical integrity index' = physinteg.1lag,
                                            'Urbanization' = urban.pct, 
                                            'Rentier state' = rentierstate,
                                            'Tax state' =  taxstate,
                                            'Winning vote-share' = winner.share.upd, 
                                            'Number of pre-election protests' = n.events.pre,
                                            'Off-schedule election' = unscheduled_election,
                                          'Regional protest diffusion' = regional_success_lag)
#incl.vars <- incl.vars %>% mutate(v2elintmon = as.numeric(v2elintmon))

@@

```{r summary-table, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Summary statistics"}
library(kableExtra)
datasummary_skim(incl.vars, output = "kableExtra")  %>%
    kable_styling(font_size = 9, full_width = FALSE)
```
\doublespace

@@

The first set of models test Hypothesis 1, on the relationship between protest risk and fraud. Protest risk is operationalized in three ways. First, the *number of pre-election protests* in ECAV is taken as *prima facie* evidence of a more turbulent and risky election environment\textemdash it indicates greater mobilization capacity in civil society and/or the opposition, and may also indicate an elevated sense of grievance. Second, the repressive capacity of the regime is modeled using the lagged *physical integrity index* for the country-year. In one approach, higher values represent more permissive environments, and should be associated with greater protest risk (and thus reduced fraud). Alternatively, in keeping with the 'murder in the middle' framework [@feinLifeIntegrityViolationsDemocracy1995], the highest protest risks (and most reduced fraud) might be found at middle values of repression; regimes with the most demonstrated repressive capacity deter potential challengers from taking to the streets in the first place, and the least repressive states defuse protest through accommodation [@pierskallaProtestDeterrenceEscalation2010;@mooreRepressionDissentSubstitution2000]. I measure repressive capacity using the *physical integrity* index from V-Dem, lagged one year. Higher values for this variable indicate greater freedom from torture and political killing. I also include the variable's square term to account for the 'murder in the middle' framework. The physical integrity variables are expected to be explanatory in models of protest, but are important control variables in the models of election fraud. Since repressive capacity helps influence protest in expectation [@Ritter2016], incumbents that are more confident of deterring protest in advance should be more willing to engage in fraud under the protest-deterrent model of manipulation.

Finally, the literature on the diffusion of electoral protest holds that successful electoral protest in one country increases the likelihood of electoral protest in similar countries nearby. This diffusion mechanism is modeled using each country's geopolitical region as coded in Teorell et al [-@qog_teorell] in conjunction with binary variables indicating a successful protest movement from NELDA. A 'successful protest' is recorded if a protest movement resulted in either a new election being run or the incumbent stepping down. The *regional diffusion* dummy variable takes on a value of one for all countries in the same region as a successful protest, for elections within one calendar year of the successfully overturned election.^[For example, the variable takes on a positive value for elections held in the post-communist region, within one year of the September 24, 2000 election in Serbia that sparked the 'Bulldozer Revolution'. All told there are 15 successful electoral protests in the dataset: India 1989, Bulgaria 1990, Bangladesh 1996, Mali 1997, Lesotho 1998, Cote d'Ivoire 2000 and 2010, Peru 2000, Serbia 2000, Georgia 2003, Ukraine 2003-2004, Azerbaijan 2005, Kyrgyzstan 2005, Haiti 2006, and Cameroon 2007.]

@@

@@

Several control variables are included, which could be common causes for both the severity of election manipulation and protest. To help control for the underlying mobilizational capacity, which likely affects incumbents' decision to manipulate and the latent risk of protest, I include lagged indicators of government *control over civil society* (the 'v2cseeorgs' variable from V-Dem), the proportion of national political parties with *permanent party organizations* (V-Dem's 'v2psorgs'), and the available of *alternative sources of information*. Since urban areas can make electoral manipulation more difficult [@ziblatt2009;@larreguy2016parties] while also making protest more likely [@Wallace2013a], I also include data on *urbanization* taken from the United Nations [@DepartmentofEconomicandSocialAffairsPopulationDivision2019]. All models include measures of *judicial constraints* and  *legislative constraints* on the executive, V-Dem variables which captures the extent to which the judicial and legislative bodies in the country act independently in practice to limit executive power. To capture the overall level of pre-election protection for rights associated with liberal democracy, I include the lagged *liberal democracy index* from V-Dem.

Finally, several binary control variables are included. First, a variable from NELDA is used to indicate *presidential* elections, since these may include both higher levels of manipulation [@Simpser2011] and a greater risk of protest. Next, a variable from NELDA indicates if the chief executive is *term limited* during the election at hand. This variable is included since incumbents running up against constitutional term limits can pose major challenges for the unity of the regime, possibly leading to fractured manipulation efforts and post-election protest [@hale_patronal_2014]. Finally, I include dummy variables that provide a categorical evaluation of state capacity. These are taken from the ordinal version of the V-Dem variable "state source of fiscal revenue" (v2stfisccap). If the V-Dem data codes a state as not capable of raising revenue or primarily relying on loans and foreign aid, I code it as a *weak state*. If the state is described as primarily relying on resource rents and expropriation, I code it is a *rentier state*. States that rely primarily on taxes on property or economic transactions are coded as *tax states*. This distinction is relevant since access to natural resource rents may both make protest less likely and repression more effective [@Girod2018]. Lastly, since strong non-democratic governments may be able to adjust the electoral calendar strategically (to affecting both election manipulation and protest risk), I include a variable from NELDA that indicates *off-schedule elections*.   Figure \@ref(fig:summary-table) shows summary statistics for all of these variables. To address concerns about multicollinearity, I present models in the appendix (page 25) that exclude control variables associated with pre-election protest.

@@

The results of these models are not promising for argument that protest risk deters election manipulation. First, Hypothesis 1 is not supported in the models shown in Table  \@ref(tab:manipdv). Models 1 through 5 apply different measures of pre-election protest risk; all of them behave contrary to the standard expectation in the literature. In Model 1, the number of pre-election protests is associated with an increase in election-day fraud; this effect is unchanged in Model 5, where *liberal democracy index* replaces other indicators of political openness. the coefficient on *pre-election protest* is positive and significant. While the effect is small, it runs counter to the expectations of the deterrence logic. A similar effect is found for *diffusion*; recent successful protest in a nearby country is also associated with a small increase in expected fraud. This suggests that the level of election manipulation, rather than being deterred by a highly contentious environment, may be deployed to make up for anticipated poor election results.

Model 2 indicates that fraud declines as societies become less repressive; an intuitive finding. However, Model 3 shows that fraud first increases with physical integrity, but then declines. In other words, fraud is lowest on the extremes of repressive capacity\textemdash where repression is severe and where it is most limited. This is contrary to the expectation that a higher risk of protest is associated with reduced fraud.

@@

```{r manipdv, echo=FALSE}
m1 <- lmer(v2elirreg.inv ~ jucon.1lag + legcon.1lag + gdpgro.1lag +  osorg.1lag  + natlpartyorg.lag   +  presidential +  urban.pct + rentierstate + taxstate + n.events.pre  +  unscheduled_election  + (1 | stateid), data = nelda2.sub)

@@

m1a <- lmer(v2elirreg.inv ~ jucon.1lag + legcon.1lag + gdpgro.1lag +  osorg.1lag  + natlpartyorg.lag   +  presidential +  urban.pct + rentierstate + taxstate +  unscheduled_election + regional_success_lag + (1 | stateid), data = nelda2.sub) 
#summary(m1a)

m1b <- lmer(v2elirreg.inv ~ jucon.1lag + legcon.1lag  + gdpgro.1lag +  osorg.1lag  + natlpartyorg.lag   + presidential   +   urban.pct + rentierstate + taxstate + physinteg.1lag +  unscheduled_election + (1 | stateid), data = nelda2.sub)

@@

m1c <- lmer(v2elirreg.inv ~ jucon.1lag + legcon.1lag + gdpgro.1lag +  osorg.1lag  + natlpartyorg.lag   +  presidential +  urban.pct + rentierstate + taxstate  + physinteg.1lag + physintlag.sq + unscheduled_election  + (1 | stateid), data = nelda2.sub) 
#summary(m1c)

m1 <- lmer(v2elirreg.inv ~ jucon.1lag + legcon.1lag + gdpgro.1lag +  osorg.1lag  + natlpartyorg.lag   +  presidential +  urban.pct + rentierstate + taxstate + n.events.pre  +  unscheduled_election  + (1 | stateid), data = nelda2.sub)

m1a <- lmer(v2elirreg.inv ~ jucon.1lag + legcon.1lag + gdpgro.1lag +  osorg.1lag  + natlpartyorg.lag   +  presidential +  urban.pct + rentierstate + taxstate +  unscheduled_election + regional_success_lag + (1 | stateid), data = nelda2.sub) 
#summary(m1a)

m1b <- lmer(v2elirreg.inv ~ jucon.1lag + legcon.1lag  + gdpgro.1lag +  osorg.1lag  + natlpartyorg.lag   + presidential   +   urban.pct + rentierstate + taxstate + physinteg.1lag +  unscheduled_election + (1 | stateid), data = nelda2.sub)

m1d <- lmer(v2elirreg.inv ~  libdem.1lag + gdpgro.1lag +    presidential +  urban.pct + rentierstate + taxstate  +  unscheduled_election + n.events.pre  + (1 | stateid), data = nelda2.sub) 
#summary(m1d)

modelsummary(c(m1, m1b, m1c, m1a, m1d), output = "kableExtra",  stars = T, gof_omit = c("R2|RMSE"), title = 'Random-intercept multilevel model of election-day fraud', coef_rename = 
               c("jucon.1lag" = "Judical independence (lag)", "legcon.1lag" = "Leg. constraints (lag)", 
                 "gdpgro.1lag" = "GDP growth rate (lag)", "altinf.1lag" = "Alternative info. (lag)", 
                 "osorg.1lag" = "Civil soc. openness (lag)", "econ.crisis.nelda.1lag1" = "Economic crisis", 
                 "natlpartyorg.lag" = "National party organization (lag)", "presidential1" = "Presidential election",
                 "inc.termlimit.num" = "Incumbent term-limited", "libdem.1lag" = "Liberal democracy index (lag)", 
                 "urban.pct" = "Urbanization", "rentierstate" = "Rentier state", "taxstate" = "Tax state",
                 "n.events.massmob.pre" = "Number of pre-election protests", "median.participants.mm.pre" = "Median size of pre-election protests", "unscheduled_election" = "Off-schedule election")) %>% kable_styling(latex_options = "HOLD_position")

@@

```{r models-hurdle-mm, warning = FALSE, message = FALSE, include = F}
model.n.novoteshare.hurd.mm <- glmmTMB::glmmTMB(n.events.post ~  jucon.1lag + legcon.1lag + altinf.1lag + libdem.1lag + gdpgro.1lag + pubseccorrup.lag + osorg.1lag  + natlpartyorg.lag + v2elirreg.inv + pubseccorrup.lag +    + v2elirreg.inv  + presidential + inc.termlimit.num  +  physinteg.1lag + physintlag.sq + urban.pct + rentierstate + taxstate + n.events.pre + unscheduled_election  + regional_success_lag,
                                ziformula=~.,
                                data = nelda2.sub %>% filter(inc.lose == "0"),
                                family=truncated_nbinom2)
summary(model.n.novoteshare.hurd.mm)

model.n.base.hurd.mm <- glmmTMB(n.events.post ~ jucon.1lag + legcon.1lag + altinf.1lag + libdem.1lag + gdpgro.1lag + pubseccorrup.lag + osorg.1lag  + natlpartyorg.lag + v2elirreg.inv + pubseccorrup.lag +    winner.share.upd + v2elirreg.inv  + presidential + inc.termlimit.num  +  physinteg.1lag + physintlag.sq + urban.pct + rentierstate + taxstate + n.events.pre + unscheduled_election + regional_success_lag,
                               ziformula=~.,
                              data = nelda2.sub %>% filter(inc.lose == "0"),
                                family=truncated_nbinom2)
summary(model.n.base.hurd.mm)

model.n.base2.hurd.mm <- glmmTMB::glmmTMB(n.events.post ~  jucon.1lag + legcon.1lag + gdpgro.1lag + altinf.1lag + libdem.1lag + pubseccorrup.lag + osorg.1lag  + natlpartyorg.lag + v2elirreg.inv + pubseccorrup.lag +    winner.share.upd*v2elirreg.inv  + presidential + inc.termlimit.num  +  physinteg.1lag + physintlag.sq + urban.pct + rentierstate + taxstate + n.events.pre + unscheduled_election + regional_success_lag,
                                          zi = ~.,
                                          family= nbinom2,
                                          data = nelda2.sub %>% filter(inc.lose == "0"))
summary(model.n.base2.hurd.mm)
